Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2024-03-22 manuscript.Rmd
add table referencing

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Marguerite Cogné, Régis Peltier, Vincent Freycon, Alexis Toumazeau, Camille Piponiot"
date: "2024-03-19"
output: officedown::rdocx_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, results = "hide")
packages_needed = c(
  "ggplot2",
  "readxl",
  "data.table",
  "knitr",
  "BIOMASS",
  "terra",
  "utils",
  "car",
  "lmerTest",
  "performance"
)

@@

lapply(packages_needed, require, character.only = TRUE)
```

@@

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural fields and diversifying agricultural landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their potential inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste:  cropping systems with fallow, silvopastoral systems, young agroforests, home gardens and forest gardens. Our results show that these traditional AFS store large amounts of carbon: the AFS with the highest carbon stocks (forest gardens) stored an average of 210\ Mg\ C\ ha$^{-1}$, similar to an old-growth tropical forest. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional agroforestry systems in Timor-Leste and their high capacity to store carbon.

@@

Agricultural expansion has been the main driver of tropical deforestation over the past half century [@Pendrill2022;@Gibbs2010;@Rudel2009], resulting in large greenhouse gas emissions [approximately 2.6 gigatonnes CO2e/yr over 2010-2014; @Pendrill2019]. The demand for tropical agricultural land is expected to continue to increase as the population grows [@UN2022]. In addition, tropical areas are expected to be disproportionately affected by climate change, putting their agriculture and food systems at risk [@Lawrence2015]. There is therefore an urgent need to promote tropical agricultural landscapes that are both more resilient to climate change, while maintaining high productivity and ecological value [@Harvey2014]. In particular, increasing the carbon stored in the soil and vegetation of agricultural landscapes could be an effective way to mitigate climate change while increasing the resilience of the food production system [@Lal2004;@Minasny2017].

@@

# for now : download manually from temporary link
if (!dir.exists("data")) create.dir("data")
# utils::unzip("data/dataverse_files.zip")
```

@@

# get plot number
data_coord[, N_PLOT := as.numeric(data.table::tstrsplit(N_Q_ID, "_")[[1]])]

@@

## Load plot general information
data_plot <- readxl::read_excel("data/DATA_BIOMASS_TL_SHORT.xlsx", "PLOT_GENERAL_INFO")

@@

```{r get-env-data}
# precipitation from CHELSA
if (!file.exists("data/CHELSA_bio12_1981-2010_V.2.1.tif")) {
  url <- "https://os.zhdk.cloud.switch.ch/envicloud/chelsa/chelsa_V2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio12_1981-2010_V.2.1.tif"
  destfile <- "data/CHELSA_bio12_1981-2010_V.2.1.tif"
  utils::download.file(url, destfile, method = "curl")

  ## crop to reduce memory usage
  box = c(124.5, 127.5, -9.2, -8.3) # Timor Leste
  rst <- terra::rast(destfile)
  rst <- terra::crop(rst, terra::ext(box))
  terra::writeRaster(rst, filename = destfile, overwrite = TRUE) 
}

# 30m resolution SRTM
if (!file.exists("data/CHELSA_bio12_1981-2010_V.2.1.tif")) {
  url <- "https://srtm.csi.cgiar.org/wp-content/uploads/files/srtm_5x5/TIFF/srtm_62_14.zip"
  destfile <- "data/srtm_62_14.zip"
  utils::download.file(url, destfile, method = "curl")
  utils::unzip(destfile, exdir = "data", overwrite = TRUE)
}

# download cwd from Chave et al 2014
if (!file.exists("data/CWD.tif")) {
  url <- "http://chave.ups-tlse.fr/pantropical_allometry/CWD.tif.zip"
  destfile <- "data/CWD.tif.zip"
  utils::download.file(url, destfile, method = "curl")
  utils::unzip(destfile, exdir = "data", overwrite = TRUE)
}
```

@@

# TODO add code to download prec and elev
site_env[, prec := terra::extract(
  terra::rast("data/CHELSA_bio12_1981-2010_V.2.1.tif"), 
  cbind(long, lat)
)]

site_env[, elev := terra::extract(
  terra::rast("data/srtm_62_14.tif"), 
  cbind(long, lat)
)]

@@

The research was carried out in the center of the Baucau district, on the north-eastern coast of Timor-Leste. Two communities, or suco (group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases: (i) Gariuai suco (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`) is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging 300-750 m; and (ii) the village of Osso Luga in Samalari suco (long-lat coordinates: `r site_summary[site=="Osso Luga", coord]`) is located on the Matebian foothill based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso Luga", prec]` mm and an altitude ranging `r site_summary[site=="Osso Luga", elev]` m (Figure\ \@ref(fig:map-sites)). The average soil properties differed between sites (Table S\@ref(tab:site-soil)), with Gariuai having Clay Loam soil texture compared to Osso Luga having clay soil texture. The two sites also have had different histories. Gariuai has been inhabited for at least 50 years, while Osso Luga was abandoned during the Indonesian occupation, and was only repopulated after 2000. During the period of abandonment, the various agroforestry systems in Osso Luga lay fallow.

@@

The AFS are presented in more detail in the supplemental information (Table S\@ref(tab:afs-typology)) and in @Cogne2024.

@@

Taxonomic identification was carried out using tetum vernacular names, which were then linked to scientific names using iNaturalist and Pl\@ntnet application, Geoffrey Hull’s monograph [@Hull2006], the data bases of “the useful tropical plants” (https://tropical.theferns.info/) as well as the native plants websites of the NT herbarium from the Australian government (https://nt.gov.au/environment/native-plants/native-plants-and-nt-herbarium). The correspondence between tetum names and scientific names is provided in the supplemental material (Table S\@ref(tab:tetum-names)).

@@

## Biomass estimation

@@

# Open large tree data (> 100 cm circumference)
data_large <- readxl::read_excel("data/DATA_BIOMASS_TL_SHORT.xlsx", "TREE INV >100CM")

# Open small tree data (30-100 cm circumference)
data_small <- readxl::read_excel("data/DATA_BIOMASS_TL_SHORT.xlsx", "TREE INV 30-100CM")

@@

# convert data_tot into data.table 
data.table::setDT(data_tot)

@@

# split scientific names into genus and species
names <- data.table::tstrsplit(data_tot$SCIENTIFIC_NAME, " ")

@@

# get wood density
data_wd <- BIOMASS::getWoodDensity(genus = data_tot$GENUS, species = data_tot$SPECIES)

# estimate mean AGC
data_tot$mean <- BIOMASS::computeAGB(
  D = data_tot$DIAM_130_CM, 
  WD = data_wd$meanWD, 
  H = data_tot$HEIGHT
) * 0.4713 # conversion factor from biomass to carbon

@@

data_sim_agc <- data.table(
  data_tot,
  BIOMASS::AGBmonteCarlo(
    D = data_tot$DIAM_130_CM, 
    WD = data_wd$meanWD, 
    errWD = data_wd$sdWD, 
    H = data_tot$HEIGHT, 
    errH = data_tot$HEIGHT*cvH, 
    Dpropag = "chave2004", 
    n = 1000,
    Carbon = TRUE
  )$AGC_simu
) |> 
  melt(
    measure.vars = c("mean", paste0("V", 1:1000)),
    variable.name = "iter",
    value.name =  "AGC"
  )

@@

We estimated tree-level AGC stocks from the measured tree CBH and height using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. The biomass of the palms was estimated by multiplying their estimated volume by their estimated wood density. The volume of the palm trunks was estimated as the volume of a cylinder, using the height and diameter measured in situ, with a volumetric shrinkage of 11% [@WD2023]. The height of the palms was measured above the leaves and was therefore multiplied by a conversion factor of 0.9 to obtain the trunk height. Biomass was then converted into carbon with a factor of 0.4713 [@Thomas2012].

@@

## Soil organic carbon estimation

@@

```{r open-soil-data}
# open soil lab data
data_soil_lab <- readxl::read_excel("data/RESUME_DATA_SOIL_SHORT_TL.xls", sheet = "RESUME-LAB ")

# open soil field measurements
data_soil_field <- readxl::read_excel("data/RAW_DATA_SOIL_LARGE_TL.xls", sheet = "SAMPLING_MEASURE")

@@

data_soil <- merge(data_soil_lab, data_soil_field, by = c("N_PLOT", "SAMPLE_DEPTH"), all=TRUE)
data.table::setDT(data_soil)
```

@@

```{r soc-error-propagation}
data_soil[, c("h0", "h1") := lapply(data.table::tstrsplit(SAMPLE_DEPTH, "-"), as.numeric)]

@@

# add total carbon stocks
data_sim = rbind(data_sim, data_sim[, .(value = sum(value[variable != "SOC10"]), variable = "TOT"), .(N_PLOT, site, ID_AF, iter)])
```

@@

## mixed effect linear regression
lmeReg <- lmerTest::lmer(log(SOC_mean) ~ log(AGC_mean) + (1|site), data = data_reg)
# Nakagawa’s R2 for mixed models
R2 = performance::r2_nakagawa(lmeReg)
summary(lmeReg)

# same for topsoil layer (0-10 cm)
lmeReg10 <- lmerTest::lmer(log(SOC10_mean) ~ log(AGC_mean) + (1|site), data = data_reg)
# Nakagawa’s R2 for mixed models
R210 = performance::r2_nakagawa(lmeReg10)
summary(lmeReg10)

@@

Total carbon stocks varied between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). The AFS with the lowest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

AGC stocks varied between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)), but not between site (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The AFS with the lowest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

SOC stocks varied between `r dataC_plot[variable=="SOC", round(min(mean))]` and `r dataC_plot[variable=="SOC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="SOC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="SOC", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). No significant differences in SOC were observed between the five AFS studied (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)). However, the measured SOC stocks differed between sites (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The SOC stocks in Gariuai were on average higher (`r dataC_site[site=="Gariuai" & variable=="SOC", round(mean)]`\ Mg\ C\ ha$^{-1}$) than those in Osso Luga (`r dataC_site[site=="Osso Luga" & variable=="SOC", round(mean)]`\ Mg\ C\ ha$^{-1}$; Figure\ \@ref(fig:barplot)).

@@

The relationship between AGC and SOC was not significant (p-value = `r round(dfreg$pvalue[grepl("30", dfreg$depth)], 2)`; Figure\ \@ref(fig:plot-lmereg)). This relationship remained insignificant (yet slightly stronger; p-value = `r round(dfreg$pvalue[grepl("10", dfreg$depth)], 2)`) when tested with SOC stocks in the topsoil layer (0-10 cm). The proportion of variance explained by the linear mixed models were `r round(dfreg$R2_cond[grepl("30", dfreg$depth)]*100)`% and `r round(dfreg$R2_cond[grepl("10", dfreg$depth)]*100)`% for 0-30 and 0-10 cm depths respectively, but a marginal R$^2$ of just `r round(dfreg$R2_marg[grepl("30", dfreg$depth)]*100, 1)`% and `r round(dfreg$R2_marg[grepl("10", dfreg$depth)]*100, 1)`% respectively when only the fixed effect was considered (Figure\ \@ref(fig:plot-lmereg)). Overall, AGC stocks were a poor predictor of SOC stocks.

@@

Compared to values reported in the literature for other AFS, the carbon stocks in this study were particularly high. The value of SOC stocks was similar to other reported values [@Feliciano2018], but some of the AFS studied here have particularly high stocks in their biomass [compared to the average of 74\ Mg\ C\ ha$^{-1}$ in @Ma2020]. The AFS with the highest carbon stocks, forest gardens, had stocks close to those of old-growth tropical forests [270\ Mg\ C\ ha$^{-1}$; @Lal2005]. These high carbon stocks could be explained by the fact that forest gardens have generally grown for more than 50 years, with little human intervention (Table S1). In terms of structure and carbon stocks, forest gardens should therefore resemble old secondary forests, which have been shown to recover more than 60% of the AGC stocks and almost all the SOC stocks of an old forest [@Poorter2021]. In addition, FGs are generally located close to villages, where soils are enriched by organic waste and animal manure, which can enhance tree growth [@Madejon2016].

@@

The location of AFS in the landscape is not random and could also explain the observed differences in AGC stocks. For example, most HGs are located close to houses, in alluvial valleys that are well supplied with water and where the soil is enriched with organic waste and animal manure, which is likely to increase soil fertility [@Palm2001]. The same applies to FGs, which are former HGs that have been more or less abandoned, and YAs, where the farmer has chosen the site with the intention of turning it into a HG. We therefore expect HGs, FGs and YAs to be located in areas that are more favorable for tree growth [@Wagner2012;@Madejon2016;@Treuter2018]. Conversely, CF and SP are mainly located on limestone plateaus where water is deep and soils are shallow, which could reduce tree growth and biomass [@Imada2008].

@@

Although agroforestry can make agriculture more resilient while improving its carbon footprint and biodiversity conservation value (Cardinael et al., 2021), poor adaptation to the local context can explain many failures to adopt new agroforestry techniques (Coe et al., 2014). Conversely, some traditional agroforestry techniques have evolved in a particular context (biophysical, social and economic) and are therefore adapted to it (Gouyon et al., 1993, Aumeeruddy-Thomas, 1994; Peltier, 1996). This is the case of the traditional AFS studied here: their heritage and socio-economic values (as shown by Cogné & Lescuyer, 2024), in addition to their high carbon stock value, confirm the importance of their conservation, both nationally and internationally..

>>>>

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Marguerite Cogné, Régis Peltier, Vincent Freycon, Alexis Toumazeau, Camille Piponiot"
date: "2024-03-19"
output:
  officedown::rdocx_document:
    number_sections: true
bibliography: references.bib
---

```{r setup, include=FALSE}
packages_needed = c(
  "ggplot2",      # version 3.4.4
  "ggrepel",      # version 0.9.4
  "ggspatial",    # version 1.1.9
  "ggpubr",       # version 0.6.0
  "readxl",       # version 1.4.3
  "data.table",   # version 1.14.8
  "knitr",        # version 1.45
  "BIOMASS",      # version 2.1.11
  "terra",        # version 1.7-55
  "car",          # version 3.1-2
  "lmerTest",     # version 3.1-3
  "performance",  # version 0.10.8
  "flextable"     # version 0.9.5
)

@@

lapply(packages_needed, require, character.only = TRUE)

opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, results = "hide")
```

@@

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural fields and diversifying agricultural landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their potential inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste:  cropping systems with fallow, silvopastoral systems, young agroforests, home gardens and forest gardens. Our results show that these traditional AFS store large amounts of carbon: the AFS with the highest carbon stocks (forest gardens) stored an average of 210\ MgC\ ha$^{-1}$, similar to an old-growth tropical forest. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional agroforestry systems in Timor-Leste and their high capacity to store carbon.

@@

Agricultural expansion has been the main driver of tropical deforestation over the past half century [@Pendrill2022;@Gibbs2010;@Rudel2009], resulting in large greenhouse gas emissions [approximately 2.6 gigatonnes CO$_2$e/yr over 2010-2014; @Pendrill2019]. The demand for tropical agricultural land is expected to continue to increase as the population grows [@UN2022]. In addition, tropical areas are expected to be disproportionately affected by climate change, putting their agriculture and food systems at risk [@Lawrence2015]. There is therefore an urgent need to promote tropical agricultural landscapes that are both more resilient to climate change, while maintaining high productivity and ecological value [@Harvey2014]. In particular, increasing the carbon stored in the soil and vegetation of agricultural landscapes could be an effective way to mitigate climate change while increasing the resilience of the food production system [@Lal2004;@Minasny2017].

@@

# for now : download manually from temporary link
if (!dir.exists("data")) create.dir("data")
unzip("data/dataverse_files.zip")
```

@@

# get plot number
data_coord[, N_PLOT := as.numeric(tstrsplit(N_Q_ID, "_")[[1]])]

@@

## Load plot general information
data_plot <- read_excel("data/DATA_BIOMASS_TL_SHORT.xlsx", "PLOT_GENERAL_INFO")

@@

```{r get-env-data}
# precipitation from CHELSA
if (!file.exists("data/CHELSA_bio12_1981-2010_V.2.1.tif")) {
  url <- "https://os.zhdk.cloud.switch.ch/envicloud/chelsa/chelsa_V2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio12_1981-2010_V.2.1.tif"
  destfile <- "data/CHELSA_bio12_1981-2010_V.2.1.tif"
  download.file(url, destfile, method = "curl")

  ## crop to reduce memory usage
  box = c(124.5, 127.5, -9.2, -8.3) # Timor Leste
  rst <- terra::rast(destfile)
  rst <- terra::crop(rst, terra::ext(box))
  writeRaster(rst, filename = destfile, overwrite = TRUE) 
}

# 30m resolution SRTM
if (!file.exists("data/CHELSA_bio12_1981-2010_V.2.1.tif")) {
  url <- "https://srtm.csi.cgiar.org/wp-content/uploads/files/srtm_5x5/TIFF/srtm_62_14.zip"
  destfile <- "data/srtm_62_14.zip"
  download.file(url, destfile, method = "curl")
  unzip(destfile, exdir = "data", overwrite = TRUE)
}

# download cwd from Chave et al 2014
if (!file.exists("data/CWD.tif")) {
  url <- "http://chave.ups-tlse.fr/pantropical_allometry/CWD.tif.zip"
  destfile <- "data/CWD.tif.zip"
  download.file(url, destfile, method = "curl")
  unzip(destfile, exdir = "data", overwrite = TRUE)
}
```

@@

# get precipitation data for the two sites
site_env[, prec := terra::extract(
  terra::rast("data/CHELSA_bio12_1981-2010_V.2.1.tif"), 
  cbind(long, lat)
)]

# get elevation data for the two sites
site_env[, elev := terra::extract(
  terra::rast("data/srtm_62_14.tif"), 
  cbind(long, lat)
)]

@@

The research was carried out in the center of the Baucau district, on the north-eastern coast of Timor-Leste. Two communities, or suco (group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases: (i) Gariuai suco (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`) is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging 300-750 m; and (ii) the village of Osso Luga in Samalari suco (long-lat coordinates: `r site_summary[site=="Osso Luga", coord]`) is located on the Matebian foothill based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso Luga", prec]` mm and an altitude ranging `r site_summary[site=="Osso Luga", elev]` m (Figure\ \@ref(fig:map-sites)). The average soil properties differed between sites (Table\ S\@ref(tab:site-soil)), with Gariuai having Clay Loam soil texture compared to Osso Luga having clay soil texture. The two sites also have had different histories. Gariuai has been inhabited for at least 50 years, while Osso Luga was abandoned during the Indonesian occupation, and was only repopulated after 2000. During the period of abandonment, the various agroforestry systems in Osso Luga lay fallow.

@@

The AFS are presented in more detail in the supplemental information (Table\ S\@ref(tab:afs-typology)) and in @Cogne2024.

@@

Taxonomic identification was carried out using tetum vernacular names, which were then linked to scientific names using iNaturalist and Pl\@ntnet application, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014] as well as the Australian Northern Territory government's Native Plants Herbarium [@NT2013]. The correspondence between tetum names and scientific names is provided in the supplemental material (Table S\@ref(tab:tetum-names)).

@@

## Carbon stocks estimation

### Biomass estimation

@@

# Open large tree data (> 100 cm circumference)
data_large <- read_excel("data/DATA_BIOMASS_TL_SHORT.xlsx", "TREE INV >100CM")

# Open small tree data (30-100 cm circumference)
data_small <- read_excel("data/DATA_BIOMASS_TL_SHORT.xlsx", "TREE INV 30-100CM")

@@

# convert data_tot into data.table 
setDT(data_tot)

@@

# split scientific names into genus and species
names <- tstrsplit(data_tot$SCIENTIFIC_NAME, " ")

@@

# get wood density
data_wd <- getWoodDensity(genus = data_tot$GENUS, species = data_tot$SPECIES)

# estimate mean AGC
data_tot$mean <- computeAGB(
  D = data_tot$DIAM_130_CM, 
  WD = data_wd$meanWD, 
  H = data_tot$HEIGHT
) * 0.4713 # conversion factor from biomass to carbon

@@

data_sim_agc <- data.table(
  data_tot,
  AGBmonteCarlo(
    D = data_tot$DIAM_130_CM, 
    WD = data_wd$meanWD, 
    errWD = data_wd$sdWD, 
    H = data_tot$HEIGHT, 
    errH = data_tot$HEIGHT*cvH, 
    Dpropag = "chave2004", 
    n = 1000,
    Carbon = TRUE
  )$AGC_simu
) |> 
  melt(
    measure.vars = c("mean", paste0("V", 1:1000)),
    variable.name = "iter",
    value.name =  "AGC"
  )

@@

We estimated tree-level AGC stocks from the measured tree CBH and height using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. The biomass of the palms was estimated by multiplying their estimated volume by their estimated wood density. The volume of the palm trunks was estimated as the volume of a cylinder, using the height and diameter measured in situ, with a volumetric shrinkage of 11% [@WD2007]. The height of the palms was measured above the leaves and was therefore multiplied by a conversion factor of 0.9 to obtain the trunk height. Biomass was then converted into carbon with a factor of 0.4713 [@Thomas2012].

@@

### Soil organic carbon estimation

@@

```{r open-soil-data}
# open soil lab data
data_soil_lab <- read_excel("data/RESUME_DATA_SOIL_SHORT_TL.xls", sheet = "RESUME-LAB ")

# open soil field measurements
data_soil_field <- read_excel("data/RAW_DATA_SOIL_LARGE_TL.xls", sheet = "SAMPLING_MEASURE")

@@

data_soil <- merge(data_soil_lab, data_soil_field, by = c("N_PLOT", "SAMPLE_DEPTH"), all=TRUE)
setDT(data_soil)
```

@@

```{r soc-error-propagation}
data_soil[, c("h0", "h1") := lapply(tstrsplit(SAMPLE_DEPTH, "-"), as.numeric)]

@@

# add total carbon stocks
data_sim = rbind(data_sim, data_sim[, .(value = sum(value[variable != "SOC10"]), variable = "TOT"), .(N_PLOT, site, ID_AF, iter)])
# group total C stocks in plots 6 and 8 (otherwise will underestimate CF total C
# stocks); this will not be used for analysis, only to display average C stocks
# per AFS
data_sim <- rbind(
  data_sim[variable != "TOT" | !N_PLOT %in% c(6, 8)], 
  data_sim[variable == "TOT" & N_PLOT %in% c(6, 8), .(value = sum(value), N_PLOT = 6), .(site, ID_AF, variable, iter)]
)
```

@@

## mixed effect linear regression
lmeReg <- lmer(log(SOC_mean) ~ log(AGC_mean) + (1|site), data = data_reg)
# Nakagawa’s R2 for mixed models
R2 = r2_nakagawa(lmeReg)
summary(lmeReg)

# same for topsoil layer (0-10 cm)
lmeReg10 <- lmer(log(SOC10_mean) ~ log(AGC_mean) + (1|site), data = data_reg)
# Nakagawa’s R2 for mixed models
R210 = r2_nakagawa(lmeReg10)
summary(lmeReg10)

@@

Total carbon stocks varied between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]`\ MgC\ ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]`\ MgC\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). The AFS with the lowest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]`\ MgC\ ha$^{-1}$; the AFS with the highest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]`\ MgC\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

AGC stocks varied between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]`\ MgC\ ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]`\ MgC\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)), but not between site (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The AFS with the lowest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]`\ MgC\ ha$^{-1}$; the AFS with the highest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]`\ MgC\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

SOC stocks varied between `r dataC_plot[variable=="SOC", round(min(mean))]` and `r dataC_plot[variable=="SOC", round(max(mean))]`\ MgC\ ha$^{-1}$, with an average of `r dataC_plot[variable=="SOC", round(mean(mean))]`\ MgC\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="SOC", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). No significant differences in SOC were observed between the five AFS studied (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)). However, the measured SOC stocks differed between sites (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The SOC stocks in Gariuai were on average higher (`r dataC_site[site=="Gariuai" & variable=="SOC", round(mean)]`\ MgC\ ha$^{-1}$) than those in Osso Luga (`r dataC_site[site=="Osso Luga" & variable=="SOC", round(mean)]`\ MgC\ ha$^{-1}$; Figure\ \@ref(fig:barplot)).

@@

The relationship between AGC and SOC was not significant (p-value = `r round(dfreg$pvalue[grepl("30", dfreg$depth)], 2)`; Figure\ \@ref(fig:plot-lmereg)). This relationship remained insignificant (p-value = `r round(dfreg$pvalue[grepl("10", dfreg$depth)], 2)`) when tested with SOC stocks in the topsoil layer (0-10 cm). The proportion of variance explained by the linear mixed models were `r round(dfreg$R2_cond[grepl("30", dfreg$depth)]*100)`% and `r round(dfreg$R2_cond[grepl("10", dfreg$depth)]*100)`% for 0-30 and 0-10 cm depths respectively, but a marginal R$^2$ of just `r round(dfreg$R2_marg[grepl("30", dfreg$depth)]*100, 1)`% and `r round(dfreg$R2_marg[grepl("10", dfreg$depth)]*100, 1)`% respectively when only the fixed effect was considered (Figure\ \@ref(fig:plot-lmereg)). Overall, AGC stocks were a poor predictor of SOC stocks.

@@

Compared to values reported in the literature for other AFS, the carbon stocks in this study were particularly high. The value of SOC stocks was similar to other reported values [@Feliciano2018], but some of the AFS studied here have particularly high stocks in their biomass [compared to the average of 74\ MgC\ ha$^{-1}$ in @Ma2020]. The AFS with the highest carbon stocks, forest gardens, had stocks close to those of old-growth tropical forests [270\ MgC\ ha$^{-1}$; @Lal2005]. These high carbon stocks could be explained by the fact that forest gardens have generally grown for more than 50 years, with little human intervention (Table S1). In terms of structure and carbon stocks, forest gardens should therefore resemble old secondary forests, which have been shown to recover more than 60% of the AGC stocks and almost all the SOC stocks of an old forest [@Poorter2021]. In addition, FGs are generally located close to villages, where soils are enriched by organic waste and animal manure, which can enhance tree growth [@Madejon2016].

@@

The location of AFS in the landscape is not random and could also explain the observed differences in AGC stocks. For example, most HGs are located close to houses, in alluvial valleys that are well supplied with water and where the soil is enriched with organic waste and animal manure, which is likely to increase soil fertility [@Palm2001]. The same applies to FGs, which are former HGs that have been more or less abandoned, and YAs, where the farmer has chosen the site with the intention of turning it into a HG. We therefore expect HGs, FGs and YAs to be located in areas that are more favorable for tree growth [@Wagner2012;@Madejon2016;@Treuer2018]. Conversely, CF and SP are mainly located on limestone plateaus where water is deep and soils are shallow, which could reduce tree growth and biomass [@Imada2008].

@@

Although agroforestry can make agriculture more resilient while improving its carbon footprint and biodiversity conservation value [@Cardinael2021], poor adaptation to the local context can explain many failures to adopt new agroforestry techniques [@Coe2014]. Conversely, some traditional agroforestry techniques have evolved in a particular context (biophysical, social and economic) and are therefore adapted to it [@Gouyon1993;@Aumeeruddy1994;@Peltier1996]. This is the case of the traditional AFS studied here: their heritage and socio-economic values [as shown by @Cogne2024], in addition to their high carbon stock value, confirm the importance of their conservation, both nationally and internationally..
